### 2022-04-06 paper/localpsf.tex
Massive cuts and edits to intro

Our motivation is Hessian approximation in distributed parameter inverse problems governed by partial differential equations (PDEs). That is, inverse problems in which one seeks to reconstruct an unknown parameter field from noisy observations of a state variable that depends on the parameter implicitly through the solution of a PDE.
In the deterministic approach to the inverse problem, one solves a regularized minimization problem to find the parameter that best fits the data. The objective function in the minimization problem is defined implicitly, so information about the Hessian of the objective function is only accessible via application of the Hessian to vectors \cite{HESSIANACTION}. 
%In particular, applying the Hessian to a vector requires solving two linear PDEs which take the form of linearizations of \eqref{eq:state_pde} \cite{HESSIANACTION}. 
%More accessible Hessian approximations are highly desirable because they permit fast solution of the minimization problem via Newton-type methods. More accessible Hessian approximations are also central to many methods for uncertainty quantification in Bayesian statistical approaches to the inverse problem, because the inverse of the Hessian locally approximates the Bayesian posterior covariance for the parameter \cite{LAPLACEAPPROXIMATION}. 
The most popular matrix-free Hessian approximation methods are based on low rank approximation \cite{CCGOPAPERS}. 
%This includes the Lanczos method, the randomized singular value decomposition \cite{HMTRANDOM}, and CUR approximation/skeletonization \cite{CUR,MAHONEY}. 
However, low rank approximation methods suffer from a ``data predicament''---if the data are highly informative about the unknown parameter, then the numerical rank of the data misfit term in the Hessian is large, so a large number of operator actions are required to form an accurate ``low rank'' approximation \cite{MYDISSERTATION}.

@@

The output of our method is a hierarchical matrix (H-matrix) approximation of a discretized version of $\Aop$. H-matrices are a matrix format in which the rows and columns of the matrix are re-ordered, then the matrix is recursively subdivided into blocks, in such a way that many off-diagonal blocks are low rank, even though the matrix as a whole may be high-rank (see Figure \ref{fig:hmatrix_intro} and Appendix \ref{app:h_matrix}). H-matrix methods permit us to perform matrix-vector products cheaply. Furthermore, one may use H-matrix methods to perform useful linear algebra operations involving the H-matrix that cannot easily be done using the original operator. These operations include matrix-matrix addition, matrix-matrix multiplication, matrix factorization, and matrix inversion. These H-matrix methods are fast and scale well to large problems. Typically, the work and memory required to perform these operations for an $N \times N$ H-matrix scales as $O(r^2 N \log(N)^a)$, where $r$ is the largest rank of the off-diagonal blocks, and $a \in \{0,1,2,3\}$. The exact cost depends on the type of H-matrix used and operation performed; for more details see \cite{HMATRIXCOST}. The ability to perform these operations permits, for example, fast solution of Newton linear systems in PDE-constrained optimization, fast sampling of ill-conditioned posterior distribution in Bayesian inverse problems, and construction of high rank surrogate models that can be used for uncertainty quantification.

@@

We overcome this difficulty using \emph{impulse response interpolation}. The impulse response, $\phi_x$, associated with a point $x$ is the Riesz representation\footnote{Recall that the Riesz representative of a functional $\rho \in L^2(\Omega)'$ with respect to the $L^2$ inner product is the unique function $\rho^* \in L^2(\Omega)$ such that $\rho(v) = \left(\rho^*,v\right)_{L^2(\Omega)}$ for all $v \in L^2(\Omega)$.} of the linear functional that results from applying $\Aop$ to a delta distribution (i.e., point source, impulse) centered at $x$. We compute batches of impulse responses by applying $\Aop$ to weighted sums of delta distributions associated with collections of points scattered throughout the domain (see Figure \ref{fig:batches_intro}). Then we interpolate shifted versions of these impulse responses to approximate arbitrary entries of the operator's integral kernel. This allows fast evaluation of approximate matrix entries, which are needed by the classical H-matrix construction methods.

>>>>

Our motivation is Hessian approximation in distributed parameter inverse problems governed by partial differential equations (PDEs). That is, inverse problems in which one seeks to reconstruct an unknown parameter field from noisy observations of a state variable that depends on the parameter implicitly through the solution of a PDE.
In the deterministic approach to the inverse problem, one solves a regularized minimization problem to find the parameter that best fits the data. The objective function in the minimization problem is defined implicitly, so information about the Hessian of the objective function is only accessible via application of the Hessian to vectors \cite{HESSIANACTION}. 
%In particular, applying the Hessian to a vector requires solving two linear PDEs which take the form of linearizations of \eqref{eq:state_pde} \cite{HESSIANACTION}. 
%More accessible Hessian approximations are highly desirable because they permit fast solution of the minimization problem via Newton-type methods. More accessible Hessian approximations are also central to many methods for uncertainty quantification in Bayesian statistical approaches to the inverse problem, because the inverse of the Hessian locally approximates the Bayesian posterior covariance for the parameter \cite{LAPLACEAPPROXIMATION}. 
The most popular matrix-free Hessian approximation methods are based on low rank approximation \cite{CCGOPAPERS}. 
%This includes the Lanczos method, the randomized singular value decomposition \cite{HMTRANDOM}, and CUR approximation/skeletonization \cite{CUR,MAHONEY}. 
However, low rank approximation methods suffer from a ``data predicament''---if the data are highly informative about the unknown parameter, then the numerical rank of the data misfit term in the Hessian is large, so a large number of operator applies are required to form an accurate ``low rank'' approximation \cite{MYDISSERTATION}. High-rank matrix free approximation methods are needed.

@@

Our method forms a hierarchical matrix (H-matrix) approximation of a discretized version of $\Aop$. H-matrices are a matrix format in which the rows and columns of the matrix are re-ordered, then the matrix is recursively subdivided into blocks, in such a way that many off-diagonal blocks are low rank, even though the matrix as a whole may be high-rank (see Figure \ref{fig:hmatrix_intro} and Appendix \ref{app:h_matrix}). H-matrix methods permit us to perform matrix-vector products cheaply. Furthermore, one may use H-matrix methods to perform useful linear algebra operations involving the H-matrix that cannot easily be done using the original operator. These operations include matrix-matrix addition, matrix-matrix multiplication, matrix factorization, and matrix inversion. These H-matrix methods are fast and scale well to large problems. Typically, the work and memory required to perform these operations for an $N \times N$ H-matrix scales as $O(r^2 N \log(N)^a)$, where $r$ is the largest rank of the off-diagonal blocks, and $a \in \{0,1,2,3\}$. The exact cost depends on the type of H-matrix used and operation performed; for more details see \cite{HMATRIXCOST}. The ability to perform these operations permits, for example, fast solution of Newton linear systems in PDE-constrained optimization, fast sampling of ill-conditioned posterior distribution in Bayesian inverse problems, and construction of high rank surrogate models that can be used for uncertainty quantification.

@@

We construct the H-matrix using \emph{impulse response interpolation}. The impulse response, $\phi_x$, associated with a point $x$ is the Riesz representation\footnote{Recall that the Riesz representative of a functional $\rho \in L^2(\Omega)'$ with respect to the $L^2$ inner product is the unique function $\rho^* \in L^2(\Omega)$ such that $\rho(v) = \left(\rho^*,v\right)_{L^2(\Omega)}$ for all $v \in L^2(\Omega)$.} of the linear functional that results from applying $\Aop$ to a delta distribution (i.e., point source, impulse) centered at $x$. We compute batches of impulse responses by applying $\Aop$ to weighted sums of delta distributions associated with collections of points scattered throughout the domain (see Figure \ref{fig:batches_intro}). Then we interpolate shifted versions of these impulse responses to approximate arbitrary entries of the operator's integral kernel. This allows fast evaluation of approximate matrix entries, which are needed by the classical H-matrix construction methods.
